Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-18 manuscript/manuscript.tex
section 3

Let $\mathbf{y}_{i}$ be the $J \times 1$ observation vector for subject $i$ such that $y_{ij}$ is the observation for subject $i$ at timepoint $j$. We assume for now that $\mathbf{y}_{i}$ is fully observed. Later, we relax this assumption by allowing for missingness in the components of $\mathbf{y}_{i}$. The general formulation of the multivariate skew normal mixture model is as follows.
\begin{equation}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k)
\end{equation}

Here $\boldsymbol\theta_k$ is the set of parameters shared by subjects belonging to cluster $k$ ($k = 1,...,K$). The probability of subject $i$ ($i = 1,...,n$) belonging to cluster $k$ ($k = 1,...,K$) is denoted $\pi_{ik}$. Note that $\pi_{ik}$ is not indexed by time and we do not allow for subjects switching clusters over the study period. This feature of the model is a result of our interest in clustering based on development patterns rather than time-specific development scores. Later, we detail a multinomial logit regression model placed on $\pi_{ik}$ to allow for flexible modeling of cluster membership probabilities as a function of covariates of interest.

In each development cluster, we assume that a subject's $J\times 1$ outcome vector follows a multivariate skew normal (MSN) distribution. We define a latent variable $Z_i \in \{1,...,K\}$ to denote the cluster membership of subject $i$ for $i = 1,...,n$. We model $\mathbf{y}_{i}|Z_i$ as 
\begin{equation}
\mathbf{y}_{i}|Z_i \sim MSN_J(\boldsymbol\zeta_k,\boldsymbol\alpha_k,\boldsymbol\Omega_k)
\end{equation}

where $Z_i = k$ is the cluster labeling of subject $i$, $\boldsymbol\zeta_k$ is a $J \times 1$ vector of cluster specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster specific scale matrix.

@@

We model the effect of covariates on longitudinal development outcomes through the use of a MSN regression model. We adopt a standard stochastic representation of a MSN random variable as the sum of a latent truncated normal random effect and a correlated multivariate normal random error term. The MSN stochastic representation is as follows.
\begin{equation}
\mathbf{y}_{i} = \boldsymbol\beta^T \mathbf{x}_{i} + t_{i}\boldsymbol\psi + \boldsymbol\epsilon_{i}
\end{equation}

@@

We note that the multivariate skew normal regression problem can be expressed more succinctly through the use of Matrix Skew Normal distributions. In particular, we define $\mathbf{Y}_{n_k \times J}$ as the response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n \mathds{1}_{(Z_i = k)}$ is the total number of observations belonging to cluster $k$ ($k = 1,...,K$). 
$$\mathbf{Y}_{n_k \times J} \sim MatSN_{n_k \times J}(\mathbf{M},\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{Y}) = \mathbf{y}_{n_kJ \times 1} = (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})$$
$$\text{vec}(\mathbf{M}) = \boldsymbol\mu_{n_k J \times 1} = (\boldsymbol\mu_1^T,...,\boldsymbol\mu_{n_k}^T)$$

where $\boldsymbol\mu_i = \mathbf{x}_i \boldsymbol\beta_k$ and $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})$. This implies that the $i^{th}$ row of $\mathbf{Y}$ is $\mathbf{y}_i \sim MSN_J(\boldsymbol\mu_i, \boldsymbol\alpha_k, \boldsymbol\Sigma_k) = {MSN}_J(\mathbf{x}_i \boldsymbol\beta_k, \boldsymbol\alpha_k, \boldsymbol\Sigma_k)$. Using the stochastic representation presented in Section 3.2, the conditional distribution of $\mathbf{y}_i | Z_i, t_i$ is $N_J (\mathbf{x}_i \boldsymbol\beta_k + \boldsymbol\psi_k t_i,\boldsymbol\Sigma_k)$. This implies that all responses for cluster $k$ conditioned on $\mathbf{T}_{n_k \times 1}$ is distributed as 
$$\mathbf{Y}_{n_k \times J} | \mathbf{T}_k = \mathbf{t}_k \sim MatNorm_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{M}_k)_{n_kJ \times 1} = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$$\

@@

A primary concern of our model is with identification of latent infant development classs. We accomplish this via multinomial logit regression model on cluster membership, which utilizes P\'{o}lya-Gamma data-augmentation, as described in Section 3.6 to allow for updating of all parameters using Gibbs sampling. The multinomial logit model is as follows for $k = 1,...,K$.
$$P(Z_i = k|\mathbf{w}_i) = \pi_{ik} = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{k' = 1}^K e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}}$$

where $\mathbf{w}_i$ is the vector of cluster probability covariates for subject $i$, $\boldsymbol\delta_k$ contains the multinomial regression parameters for cluster $k$, and $K$ is the number of putative classs specified \textit{a priori}. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $Z_i|\mathbf{w}_i \sim Multinom(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{i1},...,\pi_{1K})$. During MCMC estimation, the cluster labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as cluster assignments.

@@

We allow for missingness of outcomes in the MSN mixture model by imputing missing values from their conditional multivariate normal distributions. We note that
$$\mathbf{y}_i|Z_i,\mathbf{x}_i,t_i,\boldsymbol\beta_{k},\boldsymbol\psi_{k},\boldsymbol\Sigma_{k} \sim N_J(\boldsymbol\beta_{k}^T\mathbf{x}_i  + t_i \boldsymbol\psi_{k}, \boldsymbol\Sigma_{k})$$

where $Z_i = k$. This allows us to appeal to standard conditional forms of the multivariate normal distribution to specify the distribution of missing observations. For subject $i$, we use $q_i$ to denote the number of the total $J$ possible repeated measurements are missing. We partition the full $J \times 1$ outcome vector $\mathbf{y}_i$ into the $(J - q_i) \times 1$ observed data vector $\mathbf{y}_i^{obs}$ and the $q_i \times 1$ missing data vector $\mathbf{y}_i^{miss}$. $Y_i = [Y^{miss}_{i_{q \times 1}} | Y^{obs}_{i_{J - q \times 1}}]^T$. We have
$$\mathbf{y}_i^{miss}|\mathbf{y}_i^{obs},\mathbf{x}_i,t_i,Z_i,\boldsymbol\beta_{k},\boldsymbol\psi_{k},\boldsymbol\Sigma_{k} \sim N(\boldsymbol\mu^{miss},\boldsymbol\Sigma^{miss})$$

@@

where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = \mathds{1}_{Z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$

@@

\begin{algorithm}
\caption{Gibbs Sampler}
\begin{algorithmic}
\small
\linespread{0.5}
    \State $\mathbf{Define} \ n_{iter};\  n_{burn}; \ K; \ \boldsymbol\theta_{init}; \ \boldsymbol\theta_0$
    \State $n_{sim} := n_{iter} - n_{burn}$
    \State $\boldsymbol\theta := \boldsymbol\theta_{init}$
    \For {$\iota = 1,...,n_{sim}$}
        \State \textsc{I. Conditional Imputation}
          \For {$ i = 1,...,n$}
            \State $\mathbf{Draw} \ \mathbf{y}_i^{miss} \ \text{from} \ N_q(\boldsymbol\mu_i^{miss}, \boldsymbol\Sigma_i^{miss})$
            \State $\mathbf{y}_i := \mathbf{y}_i^{miss} \cup \mathbf{y}_i^{obs}$
         \EndFor
        \State \textsc{II. MSN Regression}
          \For {$ k = 1,...,K$}
            \State $n_k := \sum_{i = 1}^n \mathds{1}_{Z_i = k}$
            \For {$i_k = 1,...,n_k$}
              \State $\mathbf{Draw}\  t_i \ \text{from} \ N_{[0,\infty)}(a_i,A)$
            \EndFor
            \State $\mathbf{X^*}_k := $ \texttt{cbind}$(\mathbf{X}_k,\mathbf{t}_k)$
            \State $\mathbf{Draw} \ \mathbf{B^*}_k \ \text{from} \ \text{MatNorm}(\mathbf{B}_k,\mathbf{L}_k^{-1},\boldsymbol\Sigma_k)$
            \State $\mathbf{Draw} \ \boldsymbol\Sigma_k \ \text{from} \ \text{InvWish}(\nu_k, \mathbf{V}_k)$
          \EndFor
        \State \textsc{III. Multinomial Logit}
          \For {$i = 1,...,n$}
            \For {$k = 1,...,K$}
              \State $\pi_{ik} := P(Z_i = k|\mathbf{w}_i,\boldsymbol\delta_k)$
              \State $p_{ik} := P(\mathbf{y}_i|\boldsymbol\beta_k^{*T} \mathbf{x}^*_i,\boldsymbol\Sigma_k)$
            \EndFor
            \State $\mathbf{p}_{Z_i} := \frac{\mathbf{p}_i \circ \boldsymbol\pi_i}{\mathbf{p}_i \cdot \boldsymbol\pi_i}$
            \State $\mathbf{Draw} \ Z_i \ \text{from} \ \text{Categorical}(\mathbf{p}_{Z_i})$
            \For {$k = 1,...,K-1$}
              \State $\mathbf{Draw} \ \boldsymbol\delta_k \ \text{from} \ N(\mathbf{M},\mathbf{S})$
            \EndFor
          \EndFor
        \State $\boldsymbol\theta := \{\mathbf{B^*}, \boldsymbol\Sigma, \mathbf{Z}, \boldsymbol\delta \}$
        \State $\mathbf{Store} \ \boldsymbol\theta$
	  \EndFor
\end{algorithmic}
\end{algorithm}

>>>>

Let $\mathbf{y}_{i}$ be the $J \times 1$ observation vector for subject $i$ such that $y_{ij}$ is the observation for subject $i$ at timepoint $j$. We assume for now that $\mathbf{y}_{i}$ is fully observed. Later, we relax this assumption by allowing for missingness in the components of $\mathbf{y}_{i}$. For the analysis of the Nurture data, we propose the following MSN mixture
\begin{equation}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$). For now we assume that $K$ is fixed, but we discuss model selection strategies for choosing the optimal value of $K$ in Section 4. The probability that subject $i$ ($i = 1,...,n$) belongs to cluster $k$ is denoted $\pi_{ik}$. Note that $\pi_{ik}$ is indexed by $i$, as we allow the probability of class membership to vary across subjects. However, we assume a that class membership is fixed throughout the study period, since our focus is to cluster individuals based on their overall developmental patterns over the course of the study. In Section 6, we discuss extensions to allow for class membership to vary over time.

In each cluster, we assume that the $J\times 1$ outcome vector for subject $i$ follows an MSN distribution. Let $z_i \in \{1,...,K\}$ denote a latent cluster membership of subject $i$. Conditional on $z_i = k$, we model $\mathbf{y}_{i}$ as 
\begin{equation}
\mathbf{y}_{i}|z_i=k \sim MSN_J(\boldsymbol\zeta_k,\boldsymbol\alpha_k,\boldsymbol\Omega_k),
\end{equation}
where $\boldsymbol\zeta_k$ is a $J \times 1$ vector of cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that accounts for associations between the $J$ responses. The vector $\boldsymbol\alpha_k$ has components $\alpha_{jk}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate normal distribution $N_J(\boldsymbol\zeta_k,\boldsymbol\Sigma_k)$, where $\boldsymbol\Sigma_J$ is the $J \times J$ covariance matrix capturing dependence among the $J$ outcomes.

@@

We model the effect of covariates on longitudinal motor development outcomes through the use of a MSN regression model. We adopt a standard stochastic representation of a MSN random variable by conditioning on a subject-specific latent truncated normal random effect, denoted $t_i$ (Fr\"{u}hwirth-Schnatter and Pyne 2010). Conditional on membership in cluster $k$, the stochastic representation of the MSN takes the form
\begin{equation}
\mathbf{y}_{ik} = \mathbf{x}_{ik}^T\boldsymbol\beta_k  + t_{i}\boldsymbol\psi_k + \boldsymbol\epsilon_{ik}
\end{equation}

@@

We note that the multivariate skew normal regression problem can be expressed more succinctly through the use of Matrix Skew Normal distributions. In particular, we define $\mathbf{Y}_{n_k \times J}$ as the response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n \mathds{1}_{(z_i = k)}$ is the total number of observations belonging to cluster $k$ ($k = 1,...,K$). 
$$\mathbf{Y}_{n_k \times J} \sim MatSN_{n_k \times J}(\mathbf{M},\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{Y}) = \mathbf{y}_{n_kJ \times 1} = (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})$$
$$\text{vec}(\mathbf{M}) = \boldsymbol\mu_{n_k J \times 1} = (\boldsymbol\mu_1^T,...,\boldsymbol\mu_{n_k}^T)$$

where $\boldsymbol\mu_i = \mathbf{x}_i \boldsymbol\beta_k$ and $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})$. This implies that the $i^{th}$ row of $\mathbf{Y}$ is $\mathbf{y}_i \sim MSN_J(\boldsymbol\mu_i, \boldsymbol\alpha_k, \boldsymbol\Sigma_k) = {MSN}_J(\mathbf{x}_i \boldsymbol\beta_k, \boldsymbol\alpha_k, \boldsymbol\Sigma_k)$. Using the stochastic representation presented in Section 3.2, the conditional distribution of $\mathbf{y}_i | z_i, t_i$ is $N_J (\mathbf{x}_i \boldsymbol\beta_k + \boldsymbol\psi_k t_i,\boldsymbol\Sigma_k)$. This implies that all responses for cluster $k$ conditioned on $\mathbf{T}_{n_k \times 1}$ is distributed as 
$$\mathbf{Y}_{n_k \times J} | \mathbf{T}_k = \mathbf{t}_k \sim MatNorm_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{M}_k)_{n_kJ \times 1} = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$$\

@@

A primary concern of our model is with identification of latent infant development classs. We accomplish this via multinomial logit regression model on cluster membership, which utilizes P\'{o}lya-Gamma data-augmentation, as described in Section 3.6 to allow for updating of all parameters using Gibbs sampling. The multinomial logit model is as follows for $k = 1,...,K$.
$$P(z_i = k|\mathbf{w}_i) = \pi_{ik} = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{k' = 1}^K e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}}$$

where $\mathbf{w}_i$ is the vector of cluster probability covariates for subject $i$, $\boldsymbol\delta_k$ contains the multinomial regression parameters for cluster $k$, and $K$ is the number of putative classs specified \textit{a priori}. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $z_i|\mathbf{w}_i \sim Multinom(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{i1},...,\pi_{1K})$. During MCMC estimation, the cluster labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as cluster assignments.

@@

We allow for missingness of outcomes in the MSN mixture model by imputing missing values from their conditional multivariate normal distributions. We note that
$$\mathbf{y}_i|z_i,\mathbf{x}_i,t_i,\boldsymbol\beta_{k},\boldsymbol\psi_{k},\boldsymbol\Sigma_{k} \sim N_J(\boldsymbol\beta_{k}^T\mathbf{x}_i  + t_i \boldsymbol\psi_{k}, \boldsymbol\Sigma_{k})$$

where $z_i = k$. This allows us to appeal to standard conditional forms of the multivariate normal distribution to specify the distribution of missing observations. For subject $i$, we use $q_i$ to denote the number of the total $J$ possible repeated measurements are missing. We partition the full $J \times 1$ outcome vector $\mathbf{y}_i$ into the $(J - q_i) \times 1$ observed data vector $\mathbf{y}_i^{obs}$ and the $q_i \times 1$ missing data vector $\mathbf{y}_i^{miss}$. $Y_i = [Y^{miss}_{i_{q \times 1}} | Y^{obs}_{i_{J - q \times 1}}]^T$. We have
$$\mathbf{y}_i^{miss}|\mathbf{y}_i^{obs},\mathbf{x}_i,t_i,z_i,\boldsymbol\beta_{k},\boldsymbol\psi_{k},\boldsymbol\Sigma_{k} \sim N(\boldsymbol\mu^{miss},\boldsymbol\Sigma^{miss})$$

@@

where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = \mathds{1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$

@@

\begin{algorithm}
\caption{Gibbs Sampler}
\begin{algorithmic}
\small
\linespread{0.5}
    \State $\mathbf{Define} \ n_{iter};\  n_{burn}; \ K; \ \boldsymbol\theta_{init}; \ \boldsymbol\theta_0$
    \State $n_{sim} := n_{iter} - n_{burn}$
    \State $\boldsymbol\theta := \boldsymbol\theta_{init}$
    \For {$\iota = 1,...,n_{sim}$}
        \State \textsc{I. Conditional Imputation}
          \For {$ i = 1,...,n$}
            \State $\mathbf{Draw} \ \mathbf{y}_i^{miss} \ \text{from} \ N_q(\boldsymbol\mu_i^{miss}, \boldsymbol\Sigma_i^{miss})$
            \State $\mathbf{y}_i := \mathbf{y}_i^{miss} \cup \mathbf{y}_i^{obs}$
         \EndFor
        \State \textsc{II. MSN Regression}
          \For {$ k = 1,...,K$}
            \State $n_k := \sum_{i = 1}^n \mathds{1}_{z_i = k}$
            \For {$i_k = 1,...,n_k$}
              \State $\mathbf{Draw}\  t_i \ \text{from} \ N_{[0,\infty)}(a_i,A)$
            \EndFor
            \State $\mathbf{X^*}_k := $ \texttt{cbind}$(\mathbf{X}_k,\mathbf{t}_k)$
            \State $\mathbf{Draw} \ \mathbf{B^*}_k \ \text{from} \ \text{MatNorm}(\mathbf{B}_k,\mathbf{L}_k^{-1},\boldsymbol\Sigma_k)$
            \State $\mathbf{Draw} \ \boldsymbol\Sigma_k \ \text{from} \ \text{InvWish}(\nu_k, \mathbf{V}_k)$
          \EndFor
        \State \textsc{III. Multinomial Logit}
          \For {$i = 1,...,n$}
            \For {$k = 1,...,K$}
              \State $\pi_{ik} := P(z_i = k|\mathbf{w}_i,\boldsymbol\delta_k)$
              \State $p_{ik} := P(\mathbf{y}_i|\boldsymbol\beta_k^{*T} \mathbf{x}^*_i,\boldsymbol\Sigma_k)$
            \EndFor
            \State $\mathbf{p}_{z_i} := \frac{\mathbf{p}_i \circ \boldsymbol\pi_i}{\mathbf{p}_i \cdot \boldsymbol\pi_i}$
            \State $\mathbf{Draw} \ z_i \ \text{from} \ \text{Categorical}(\mathbf{p}_{z_i})$
            \For {$k = 1,...,K-1$}
              \State $\mathbf{Draw} \ \boldsymbol\delta_k \ \text{from} \ N(\mathbf{M},\mathbf{S})$
            \EndFor
          \EndFor
        \State $\boldsymbol\theta := \{\mathbf{B^*}, \boldsymbol\Sigma, \mathbf{Z}, \boldsymbol\delta \}$
        \State $\mathbf{Store} \ \boldsymbol\theta$
	  \EndFor
\end{algorithmic}
\end{algorithm}
